---
title: "Basic SQL Tutorial Notes"
output: html_document
---

# Basic SQL Tutorial Notes

As a way to practice SQL and tidyverse at the same time, I'll draw parallels between commands from the Mode SQL tutorial and the equivalent command using dplyr.

## SELECT

### Select specific columns from table

```{sql}
SELECT year,
       month,
       west
  FROM tutorial.us_housing_units
```

```{r}
# dplyr
tutorial.us_housing_units %>% select (year,month,west)
# base r
tutorial.us_housing_units[,c("year","month","west")]

```

Note that there is no comma after "west", the last column selected from tutorial.us_housing_units. If a comma is put after west, the command will error. FROM must come after SELECT. Also, SELECT and FROM don't need to be capitalized but the convention is to capitalize them. The indentation is also optional, but makes things easier to read.

### Select all columns from table

```{sql}
SELECT *
  FROM tutorial.us_housing_units

--equivalent SQL command
SELECT year,
       month,
       month_name,
       west,
       midwest,
       south,
       northeast
  FROM tutorial.us_housing_units
```

```{r}
# just running the name of table itself will return everything
tutorial.us_housing_units

# or manually select all columns
tutorial.us_housing_units %>% select (year,month,month_name,west,midwest,south,northeast)
```

### Rename columns from a selection

```{sql}
SELECT west AS West_Region,
       south AS South_Region
  FROM tutorial.us_housing_units
```

Note - if not put in double quotes, the command above will return columns all lowercase (e.g. west_region).

```{r}
# use the select and rename functions in dplyr
tutorial.us_housing_units %>% 
  select(west,south) %>%
  rename ("west_region" = west,
          "south_region" = south)
```

```{sql}
SELECT year AS "Year",
       month AS "Month",
       month_name AS "Month Name",
       west AS "West",
       midwest AS "Midwest",
       south AS "South",
       northeast AS "Northeast"
  FROM tutorial.us_housing_units
```

```{r}

tutorial.us_housing_units %>% rename ("Year" = year, 
                                      "Month" = month,
                                      "Month Name" = month_name,
                                      "West" = west,
                                      "Midwest"=midwest,
                                      "South" = south,
                                      "Northeast" = northeast)

```

Note - single quotes are different than double quotes - if you use single quotes for year AS 'Year', you will get an error.


## LIMIT

# Apply a limit to only return first 15 rows of data

```{sql}
SELECT *
  FROM tutorial.us_housing_units
 LIMIT 15
```

```{r}
# use the head function
tutorial.us_housing_units %>% head(15)
```

Important note - LIMIT must come after FROM, or else it will error.

## WHERE

```{sql}
SELECT *
  FROM tutorial.us_housing_units
 WHERE month = 1
```

The clauses always need to be in this order: SELECT, FROM, WHERE.
Be careful with distinction between R / SQL in which SQL uses just one '=' and R requires '=='

```{r}
tutorial.us_housing_units %>% filter (month == 1)
```

Note - again, single quotes are different than double quotes - if you use single quotes for the character, such as writing WHERE month_name > "J", this will error.

## Did the West Region ever produce more than 50,000 housing units in one month?

```{sql}
SELECT *
  FROM tutorial.us_housing_units
  WHERE west > 50
```

Running this query returns 3 rows where the west region has produced more than 50,000 units (greater than value of 50 in the table).

```{r}
# this will evaluate to TRUE if there are any records with more than 50000 units in west region
# calculates the number of rows after filtering the dataset, and tests whether it is greater than 0
nrow(tutorial.us_housing_units %>% filter(west>50)) > 0
```

## Did the South Region ever produce 20,000 or fewer housing units in one month?

```{sql}
SELECT *
  FROM tutorial.us_housing_units
  WHERE south <= 20
```

There were 4 months where the south region produced less than 20,000 units.

## Write a query that only shows rows for which the month name is February.

```{r}
SELECT *
  FROM tutorial.us_housing_units
  WHERE month_name = 'February'
```

## Write a query that only shows rows for which the month_name starts with the letter "N" or an earlier letter in the alphabet.

```{r}
SELECT *
  FROM tutorial.us_housing_units
  WHERE month_name < 'o'
```


## Arithmetic on columns

```{r}
SELECT year,
       month,
       west,
       south,
       west + south AS south_plus_west
  FROM tutorial.us_housing_units
```

```{r}
tutorial.us_housing_units %>% select (year,month,west,south) %>% mutate (south_plus_west = west + south)
```

## Write a query that calculates the sum of all four regions in a separate column.

```{r}
SELECT year,
       month,
       west,
       south,
       northeast,
       midwest,
       west + south + northeast + midwest AS sum_territories
  FROM tutorial.us_housing_units
  
```

## Write a query that returns all rows for which more units were produced in the West region than in the Midwest and Northeast combined.

```{r}
SELECT *
  FROM tutorial.us_housing_units
  WHERE west > (midwest +northeast)
```

Tried to make a new column and then filter on it, but didn't work (don't do the below):

```{r}
SELECT *,
       midwest+northeast AS mid_and_north
  FROM tutorial.us_housing_units
  WHERE west > mid_and_north
```


## Write a query that calculates the percentage of all houses completed in the United States represented by each region. Only return results from the year 2000 and later.

Hint: There should be four columns of percentages.

```{sql}
SELECT *,
       west/(west+south+midwest+northeast)*100 AS west_percent,
       south/(west+south+midwest+northeast)*100 AS south_percent,
       northeast/(west+south+midwest+northeast)*100 AS northeast_percent,
       midwest/(west+south+midwest+northeast)*100 AS midwest_percent
  FROM tutorial.us_housing_units
  WHERE year >= 2000
```





