---
title: "Basic SQL Tutorial Notes"
output: html_document
---

# Basic SQL Tutorial Notes

As a way to practice SQL and tidyverse at the same time, I'll draw parallels between commands from the Mode SQL tutorial and the equivalent command using dplyr.

## SELECT

### Select specific columns from table

```{sql}
SELECT year,
       month,
       west
  FROM tutorial.us_housing_units
```

```{r}
# dplyr
tutorial.us_housing_units %>% select (year,month,west)
# base r
tutorial.us_housing_units[,c("year","month","west")]

```

Note that there is no comma after "west", the last column selected from tutorial.us_housing_units. If a comma is put after west, the command will error. FROM must come after SELECT. Also, SELECT and FROM don't need to be capitalized but the convention is to capitalize them. The indentation is also optional, but makes things easier to read.

### Select all columns from table

```{sql}
SELECT *
  FROM tutorial.us_housing_units

--equivalent SQL command
SELECT year,
       month,
       month_name,
       west,
       midwest,
       south,
       northeast
  FROM tutorial.us_housing_units
```

```{r}
# just running the name of table itself will return everything
tutorial.us_housing_units

# or manually select all columns
tutorial.us_housing_units %>% select (year,month,month_name,west,midwest,south,northeast)
```

### Rename columns from a selection

```{sql}
SELECT west AS West_Region,
       south AS South_Region
  FROM tutorial.us_housing_units
```

Note - if not put in double quotes, the command above will return columns all lowercase (e.g. west_region).

```{r}
# use the select and rename functions in dplyr
tutorial.us_housing_units %>% 
  select(west,south) %>%
  rename ("west_region" = west,
          "south_region" = south)
```

```{sql}
SELECT year AS "Year",
       month AS "Month",
       month_name AS "Month Name",
       west AS "West",
       midwest AS "Midwest",
       south AS "South",
       northeast AS "Northeast"
  FROM tutorial.us_housing_units
```

```{r}

tutorial.us_housing_units %>% rename ("Year" = year, 
                                      "Month" = month,
                                      "Month Name" = month_name,
                                      "West" = west,
                                      "Midwest"=midwest,
                                      "South" = south,
                                      "Northeast" = northeast)

```

## LIMIT

# Apply a limit to only return first 15 rows of data

```{sql}
SELECT *
  FROM tutorial.us_housing_units
 LIMIT 15
```

```{r}
# use the head function
tutorial.us_housing_units %>% head(15)
```

Important note - LIMIT must come after FROM, or else it will error.

## WHERE

```{sql}
SELECT *
  FROM tutorial.us_housing_units
 WHERE month = 1
```

Important note - the clauses always need to be in this order: SELECT, FROM, WHERE.
Be careful with distinction between R / SQL in which SQL uses just one '=' and R requires '=='

```{r}
# be careful 
tutorial.us_housing_units %>% filter (month == 1)
```




